---
title: "GoM_manuscript_graphics"
author: "WHMackenzie"
date: "`r Sys.Date()`"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, cache = FALSE)
require(mapview)
require(rasterVis) 
require(ggplot2)
require(sf)
require(fasterize)
require(purrr)
require(parallel)
require(tmap)
require(tidyverse)
require(data.table)
require(DBI)
require(climr)
require(tictoc)
require(ggdensity)
library(gt)
library(cowplot)
```

```{r}
tree.codes <- fread("./clean_data/Tree_Species_Codes.csv")
```
##Summary of GoM trials
```{r}

gom.trials <- fread("./clean_data/gom_trials_compiled.csv")
# Step 1: Create the main summary table
gom.trials.summary <- gom.trials %>%
  group_by(species, year_established) %>%
  tally(name = "n") %>%
  pivot_wider(
    names_from = year_established,
    values_from = n,
    values_fill = 0
  ) %>%
  ungroup()
# Summarize total trees planted per species per year

# Step 2: Reorder year columns numerically
year_cols <- sort(setdiff(names(gom.trials.summary), "species"))


gom.trials.summary <- gom.trials.summary %>%
  dplyr::select(species, all_of(year_cols)) %>%
  mutate(Total_Trials = rowSums(across(all_of(year_cols)))) %>%
  relocate(Total_Trials, .after = last_col()) %>%
  arrange(species)

# Step 3: Create a summary row for column totals
column_totals <- gom.trials.summary %>%
  summarise(across(all_of(year_cols), sum)) %>%
  mutate(
    species = "Total",
    Total_Trials = sum(across(all_of(year_cols)))
  ) %>%
  dplyr::select(names(gom.trials.summary))  # ensure column order matches

# Step 4: Bind the summary row to the bottom
gom.trials.summary <- bind_rows(gom.trials.summary, column_totals)

# Step 5: Render with gt
gt::gt(gom.trials.summary) %>%
  tab_header(
  title = md(""),
  subtitle = md("")
  ) %>%
  fmt_number(decimals = 0) %>%
  cols_label(
    species = "Species",
    Total_Trials = "Total Trials"
  ) %>%
  #opt_align_table_header("left") %>%
  tab_options(
    table.font.size = px(10),
    #heading.title.font.size = px(10),
    #heading.subtitle.font.size = px(10),
    column_labels.font.size = px(10),
    data_row.padding = px(2)
  ) %>%
  gtsave(
    filename = "./graphics/GoM_trial_summary_by_species_year2.png",
    vwidth = 635,   # 16.8 cm
    vheight = 600
  )

#fwrite(gom.trials.summary, "./clean_data/GoM_trial_summary_by_species_year.csv")
```



## Geographic distances from GoM trials to nearest natural stand and plot
```{r}
library(sf)
library(dplyr)
library(data.table)
library(geosphere)  # For geodesic distance calculation

# Load and prepare spatial data (already done in your script)
tree.climate2 <- fread("./clean_data/Tree_Plot_Climate.csv")
tree.sf <- st_as_sf(tree.climate2, coords = c("Longitude", "Latitude"), crs = 4326) %>%
  left_join(tree.codes, by = c("Species" = "Code"))
gom.climate <- fread("./clean_data/GoM_trial_climate.csv") %>%
  filter(species %in% c("Cw","Fd","Lw","Pw","Py"))
gom.sf <- st_as_sf(gom.climate, coords = c("longitude", "latitude"), crs = 4326) %>%
  filter(species %in% c("Cw", "Fd", "Lw", "Pw", "Py"))

gom.assessed <- fread("./clean_data/GoM_assessed_climate.csv")
gom.assessed.sf <- st_as_sf(gom.assessed, coords = c("longitude", "latitude"), crs = 4326)

# Function to calculate nearest distance (in km) from each trial to natural stand
calculate_nearest_distance <- function(trial_sf, natural_sf) {
  trial_coords <- st_coordinates(trial_sf)
  natural_coords <- st_coordinates(natural_sf)

  nearest_distances <- sapply(1:nrow(trial_coords), function(i) {
    trial_point <- trial_coords[i, ]
    dists <- distHaversine(trial_point, natural_coords) / 1000  # Convert to km
    min(dists)
  })

  trial_sf$nearest_natural_km <- nearest_distances
  return(trial_sf)
}

# Loop over species and calculate distances
distance_results <- list()
assessed_distance <- list()
for (sp in unique(gom.sf$species)) {
  tree_data <- tree.sf %>% filter(species == sp)
  trial_data <- gom.sf %>% filter(species == sp)
  if (nrow(tree_data) > 0 && nrow(trial_data) > 0) {
    trial_with_dist <- calculate_nearest_distance(trial_data, tree_data)
    distance_results[[sp]] <- trial_with_dist
  }
}

for (sp in unique(gom.assessed.sf$species)) {
  tree_data <- tree.sf %>% filter(species == sp)
  assessed_data <- gom.assessed.sf %>% filter(species == sp)
  if (nrow(tree_data) > 0 && nrow(trial_data) > 0) {
    assessed_with_dist <- calculate_nearest_distance(assessed_data, tree_data)
    assessed_distance[[sp]] <- assessed_with_dist
  }
}


# Combine all species results into one sf object
all_gom_trials <- do.call(rbind, distance_results)
assessed_gom_trials <- do.call(rbind, assessed_distance)  %>%  distinct
# Merge assessed distances back to main GoM trials data
# gom_trials <- left_join(st_drop_geometry(all_gom_trials), st_drop_geometry(assessed_gom_trials), by = "PlotNumber") 
# # Optional: Save to CSV or shapefile
# # st_write(gom_trials_with_distances, "./output/gom_trials_with_distances.gpkg")
# fwrite(gom_trials, "./clean_data/all_gom_trials_data.csv")

library(ggplot2)
library(dplyr)

# Ensure 'species' and 'nearest_natural_km' are present
summary_data <- all_gom_trials %>%
  #st_drop_geometry() %>%
  filter(!is.na(nearest_natural_km)) %>%
  mutate(species = factor(species))  # Optional: reorder by median distance

# Plot: Boxplot with jittered points
# ggplot(summary_data, aes(x = species, y = nearest_natural_km, fill = species)) +
#   geom_boxplot(outlier.shape = NA, alpha = 0.6) +
#   geom_jitter(width = 0.2, alpha = 0.5, color = "black", size = 1.5) +
#   scale_y_continuous(name = "Nearest Natural Stand (km)") +
#   scale_x_discrete(name = "Species") +
#   theme_minimal(base_size = 14) +
#   theme(legend.position = "none") #+
#   #labs(
#     #title = "Distance from GoM Trials to Nearest Natural Plot",
#     #subtitle = "By Species",
#     #caption = "Distances calculated using geodesic (Haversine) method"
#   #)

distance_plot <- ggplot(summary_data, aes(x = species, y = nearest_natural_km, fill = species)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.2, alpha = 0.5, color = "black", size = 1) +
  scale_y_continuous(name = "Km") +
  scale_x_discrete(NULL)+#name = "Species") +
  theme_minimal(base_size = 10) +
  # theme(
  #   legend.position = "none",
  #   panel.border = element_rect(color = "black", fill = NA, linewidth = 0.6),
  #   plot.background = element_rect(color = "black", fill = NA, linewidth = 0.6)
  # ) +
    theme(
    axis.title.x = element_text(size = 8),
    axis.title.y = element_text(size = 8),
    axis.text = element_text(size = 8),
    axis.ticks = element_line(color = "grey70", linewidth = 0.2),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.6),
    plot.background = element_rect(color = "black", fill = NA, linewidth = 0.6),
    plot.title = element_text(size = 10, face = "bold")
  )+
  labs(title = "Distance to Nearest Natural Stand")
 distance_plot #
ggsave("./graphics/GoM_trial_distances_to_natural_plots.png", plot = distance_plot, width = 8.6, height = 6, units = "cm")


```

## Choose dimensions of prov map to use
```{r}
tree.climate2 <- fread("./clean_data/Tree_Plot_Climate.csv")
tree.sf <- st_as_sf(tree.climate2, coords = c("Longitude", "Latitude"), crs = 4326) %>%
  left_join(tree.codes, by = c("Species" = "Code"))

gom.sf <- st_as_sf(gom.climate, coords = c("longitude", "latitude"), crs = 4326) %>%
  filter(species %in% c("Cw", "Fd", "Lw", "Pw", "Py"))

provmap <- st_read("D:/OneDrive - Government of BC/CommonTables/BC_AB_US_Shp/BC_Province_Outline_Dissolved.gpkg")
bbox <- st_bbox(provmap)
fix_bbox <- function(b) {
  b <- as.numeric(b)
  names(b) <- c("xmin", "ymin", "xmax", "ymax")
  class(b) <- c("bbox")
  attr(b, "crs") <- NA
  b
}

shrink_bbox_asymmetric <- function(bbox,
                                   shrink_left   = 0,
                                   shrink_right  = 0,
                                   shrink_bottom = 0,
                                   shrink_top    = 0) {
  
  # Ensure bbox is numeric and named
  b <- as.numeric(bbox)
  names(b) <- c("xmin", "ymin", "xmax", "ymax")
  
  # Extract CRS if present
  crs <- attr(bbox, "crs")
  
  width  <- b["xmax"] - b["xmin"]
  height <- b["ymax"] - b["ymin"]
  
  new <- c(
    xmin = b["xmin"] + width  * shrink_left,
    ymin = b["ymin"] + height * shrink_bottom,
    xmax = b["xmax"] - width  * shrink_right,
    ymax = b["ymax"] - height * shrink_top
  )
  
  # Rebuild as a proper bbox object
  class(new) <- "bbox"
  attr(new, "crs") <- crs
  
  new
}

bbox_new <- shrink_bbox_asymmetric(
  bbox,
  shrink_top = .2,      # remove 30% from top
  shrink_bottom = .05,# remove 10% from bottom
  shrink_left = .25,
  shrink_right = 0.05
)
provmap_crop <- st_crop(provmap, bbox_new)

ggplot() +
  geom_sf(data = provmap_crop) +
coord_sf(
  xlim = c(bbox_new["xmin"], bbox_new["xmax"]),
  ylim = c(bbox_new["ymin"], bbox_new["ymax"]),
  expand = FALSE,
  crs = st_crs(provmap)
) +
  theme_bw()

#provmap_crop = provmap
```

#Graphic with maps of tree plot locations and GoM trials 
```{r}
library(data.table)
library(dplyr)
library(sf)
library(ggplot2)
library(cowplot)

#------------------------------------------------------------
# 1. Load assessed GoM trials
#------------------------------------------------------------
gom.assessed <- fread("./clean_data/GoM_assessed_trials_2023_24.csv")

gom.assessed.sf <- st_as_sf(
  gom.assessed,
  coords = c("Block_long", "Block_lat"),
  crs = 4326
)

#------------------------------------------------------------
# 2. Define performance categories and species names
#------------------------------------------------------------
performance_levels <- c("Fail", "Poor", "Fair", "Good", "Excellent")

performance_colors <- c(
  "Excellent" = "purple",
  "Good"      = "green",
  "Fair"      = "yellow",
  "Poor"      = "orange",
  "Fail"      = "red"
)

species_names <- c(
  "Cw" = "Cw - Western redcedar",
  "Fd" = "Fd - Douglas-fir",
  "Lw" = "Lw - Western larch",
  "Pw" = "Pw - Western white pine",
  "Py" = "Py - Ponderosa pine"
)

#------------------------------------------------------------
# 3. Build species‑specific maps
#------------------------------------------------------------
plots <- list()

for (sp in unique(gom.sf$species)) {

  gom_data <- gom.sf %>%
    filter(species == sp) %>%
    mutate(type = "GoM trial")

  tree_data <- tree.sf %>%
    filter(species == sp) %>%
    mutate(type = "Natural stand")

  assessed_data <- gom.assessed.sf %>%
    filter(species == sp) %>%
    mutate(performance = factor(performance, levels = performance_levels))

  #--------------------------------------------------------
  # Jitter assessed points to avoid overplotting
  #--------------------------------------------------------
  jittered_coords <- st_coordinates(assessed_data) +
    matrix(runif(nrow(assessed_data) * 2, -0.03, 0.03), ncol = 2)

  assessed_data_jittered <- st_as_sf(
    cbind(assessed_data, jittered_coords),
    coords = c("X", "Y"),
    crs = 4326
  )

  #--------------------------------------------------------
  # Build the species map
  #--------------------------------------------------------
  p <- ggplot() +
    geom_sf(data = provmap_crop, fill = "grey90", color = "black") +

    geom_sf(data = gom_data,
            aes(shape = type),
            color = "black", size = 2, alpha = 0.8) +

    geom_sf(data = assessed_data_jittered,
            aes(color = performance),
            size = 2, shape = 17, alpha = 0.9) +

    geom_sf(data = tree_data,
            aes(shape = type),
            color = "darkblue", size = 0.8, alpha = 0.8) +

    scale_shape_manual(values = c("Natural stand" = 1, "GoM trial" = 2)) +
    scale_color_manual(values = performance_colors,
                       breaks = performance_levels) +

    guides(
      shape = guide_legend(order = 1),
      color = guide_legend(order = 2)
    ) +

    coord_sf(
      xlim = st_bbox(provmap_crop)[c("xmin", "xmax")],
      ylim = st_bbox(provmap_crop)[c("ymin", "ymax")],
      expand = FALSE
    ) +

theme_minimal() +
theme(
  panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8),
  plot.background = element_rect(fill = NA, color = NA),
  plot.margin = margin(0, 0, 0, 0),

  axis.title.x = element_text(size = 8),
  axis.title.y = element_text(size = 8),
  axis.text = element_text(size = 8),
  axis.ticks = element_line(color = "grey70", linewidth = 0.2)
) +

    labs(
      title = species_names[sp],
      shape = "Source",
      color = "GoM Assessed"
    )

  plots[[sp]] <- p
}

#------------------------------------------------------------
# 4. Reorder species panels
#------------------------------------------------------------
species_order <- c("Cw", "Fd", "Lw", "Pw", "Py")
plots_ordered <- plots[species_order]

#------------------------------------------------------------
# 5. Extract shared legend
#------------------------------------------------------------
legend <- cowplot::get_legend(
  plots_ordered[[1]] +
    scale_shape_manual(values = c("Natural stand" = 1, "GoM trial" = 2)) +
    scale_color_manual(values = performance_colors,
                       breaks = performance_levels) +
    theme(
      legend.position = "bottom",
      legend.box = "horizontal",
      legend.justification = "left",
      legend.box.just = "left",
      legend.text = element_text(hjust = 0),
      legend.title = element_text(hjust = 0)
    )
)

#------------------------------------------------------------
# 6. Add distance plot and remove legends from all panels
#------------------------------------------------------------
distance_plot2 <- distance_plot + 
  #coord_fixed(ratio = .5)+   # adjust ratio if needed
theme_minimal() +
theme(
  panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8),
  plot.background = element_rect(fill = NA, color = NA),
  plot.margin = margin(0, 0, 0, 0),

  axis.title.x = element_text(size = 8),
  axis.title.y = element_text(size = 8),
  axis.text = element_text(size = 8),
  axis.ticks = element_line(color = "grey70", linewidth = 0.2)
)



plots_ordered_with_dist <- c(plots_ordered, list(distance_plot2))

plots_nolegend <- lapply(
  plots_ordered_with_dist,
  function(x) x + theme(legend.position = "none")
)

#------------------------------------------------------------
# 7. Combine into 2 rows (Lw, Pw, Py) / (Cw, Fd)
#------------------------------------------------------------
combined_map_plot <- plot_grid(
  plotlist = plots_nolegend,
  nrow = 2,
  labels = "AUTO",
  label_size = 10,
  label_fontface = "bold",
  label_x = 0.03,
  label_y = 0.95,
  #rel_heights = c(1, 1),
  hjust = 0,
  vjust = 0.5,
  align = "hv",
  axis = "tblr"
)
combined_map_plot
# Save intermediate version
ggsave("./graphics/all_species_maps_new.jpg",
       combined_map_plot,
       width = 24.4, height = 16, units = "cm", dpi = 600)

#------------------------------------------------------------
# 8. Add shared legend below
#------------------------------------------------------------
final_plot <- plot_grid(
  combined_map_plot,
  legend,
  ncol = 1,
  rel_heights = c(1, 0.12))#,
#   align = "hv",
#   axis = "lr"
# )

final_plot

ggsave("./graphics/all_species_maps_final.jpg",
       final_plot,
       width = 24.4, height = 16, units = "cm", dpi = 600)
```

Assessed GoM trial performance
```{r}
#assessed <- fread( "./clean_data/GoM_assessed_trials_2023_24.csv")
gom.assessed <- fread("./clean_data/GoM_assessed_climate.csv") %>%
  filter(species %in% c("Cw","Fd","Lw","Pw","Py"),
         PERIOD == "1961_1990") %>%
  arrange(species) %>%
  mutate(performance = factor(performance,
                              levels = c("Fail", "Poor", "Fair", "Good", "Excellent"))) %>%
  distinct() %>%
  drop_na(performance)
# Prepare data with counts and percentages
performance_summary <- gom.assessed %>% filter(species %in% c("Cw", "Fd", "Lw", "Pw", "Py")) %>% drop_na %>% 
  count(species, performance) %>% #, Planted_year) %>%
  group_by(species) %>% #, Planted_year) %>%
  mutate(percent = n / sum(n) * 100) %>%
  ungroup() %>%
  mutate(performance = factor(performance,
                              levels = c("Excellent", "Good", "Fair", "Poor", "Fail"),
                              ordered = TRUE))

total_counts <- performance_summary %>%
  #group_by(Planted_year) %>%
  summarise(total_n = sum(n), .groups = 'drop')
# Stacked bar plot with counts and percentage labels
ggplot(performance_summary, aes(x = species, y = n, fill = performance)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(aes(label = paste0(round(percent), "%")),
            position = position_stack(vjust = 0.5),
            size = 3, color = "black") +
  theme_minimal() +
  labs(#title = "Performance Distribution by Species in GoM Assessed Trials planted 2023-24",
       #subtitle = "242 trials assessed across 5 species",
       x = "Species",
       y = "Count of Trials",
       fill = "Performance") +
  scale_fill_manual(values = c("Excellent" = "purple",
                               "Good" = "green",
                               "Fair" = "yellow",
                               "Poor" = "orange",
                               "Fail" = "red")) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(size = 10, face = "bold"),
        strip.text = element_text(size = 10, face = "bold"))#+
 # facet_wrap(~Planted_year)

ggsave("./graphics/GoM_assessed_trials_performance_2023_24.png", width = 7.3725, height = 4)
```
create hdr plots by species
```{r}
make_hdr_panel <- function(sp, clim_x, clim_y) {

  ggplot() +
    # HDR contours
    geom_hdr(
      data = tree.climate2 %>% filter(species == sp),
      aes(x = .data[[clim_x]], y = .data[[clim_y]]),
      probs = c(0.90, 0.75, 0.5)
    ) +

    # All GoM trials
    geom_point(
      data = gom.climate %>% filter(species == sp),
      aes(x = .data[[clim_x]], y = .data[[clim_y]]),
      color = "grey20", alpha = 0.3, size = 1
    ) +

    # Assessed trials
    geom_point(
      data = gom.assessed %>% filter(species == sp),
      aes(x = .data[[clim_x]], y = .data[[clim_y]], color = performance),
      size = 2, shape = 17, alpha = 0.9
    ) +

    scale_color_manual(values = performance_colors, drop = FALSE) +

    # Axis limits (still hard‑coded; can parameterize if you want)
    coord_cartesian(xlim = c(250, 2000), clip = "off") +
    scale_x_continuous(limits = c(250, 2000)) +

    theme_minimal(base_size = 10) +
    theme(
      panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5),
      plot.background = element_rect(fill = NA, color = NA),
      plot.margin = margin(2, 2, 2, 2),
      legend.position = "none",
      axis.title.x = element_text(size = 8),
      axis.title.y = element_text(size = 8)
    ) +

    labs(
      title = species_names[sp],
      x = clim_x,
      y = clim_y
    )
}

species_order <- c("Cw","Fd","Lw","Pw","Py")
clim_x = "DD5_an"
clim_y = "MCMT"
hdr_panels <- lapply(
  species_order,
  function(sp) make_hdr_panel(sp, clim_x, clim_y)
)

names(hdr_panels) <- species_order
```

```{r}
library(data.table)
library(dplyr)
library(ggplot2)
library(ggdensity)
library(patchwork)

# -----------------------------
# 1. Select climate variables
# -----------------------------
clim_x <- "DD5_an"
clim_y <- "MCMT"

# -----------------------------
# 2. Load and filter data
# -----------------------------
tree.climate2 <- fread("./clean_data/Tree_Plot_Climate.csv") %>%
  filter(species %in% c("Cw","Fd","Lw","Pw","Py")) %>%
  distinct()

gom.climate <- fread("./clean_data/GoM_trial_climate.csv") %>%
  filter(species %in% c("Cw","Fd","Lw","Pw","Py"))

WNA.climate <- fread("D:/OneDrive - Government of BC/CommonTables/BGC_climate/WNA_BGCv13_10km.csv") %>%
  filter(!is.na(MAT), !is.na(MAP), !is.na(MSP), !is.na(MWMT),
         !is.na(CMD_an), !is.na(DD5_an))

gom.assessed <- fread("./clean_data/GoM_assessed_climate.csv") %>%
  filter(species %in% c("Cw","Fd","Lw","Pw","Py"),
         PERIOD == "1961_1990") %>%
  arrange(species) %>%
  mutate(performance = factor(performance,
                              levels = c("Fail", "Poor", "Fair", "Good", "Excellent"))) %>%
  distinct() %>%
  drop_na(performance)

# -----------------------------
# 3. Species names & colors
# -----------------------------
performance_colors <- c(
  "Fail" = "red",
  "Poor" = "orange",
  "Fair" = "yellow",
  "Good" = "green",
  "Excellent" = "purple"
)

species_names <- c(
  "Cw" = "Western redcedar",
  "Fd" = "Douglas-fir",
  "Lw" = "Western larch",
  "Pw" = "Western white pine",
  "Py" = "Ponderosa pine"
)

tree.climate2[, species_full := species_names[species]]
gom.assessed[, species_full := species_names[species]]
gom.climate[, species_full := species_names[species]]

# -----------------------------
# 4. HDR Climate Density Plot
# -----------------------------
hdr.plot <- ggplot(tree.climate2,
                   aes(x = .data[[clim_x]], y = .data[[clim_y]])) +

  geom_hdr(
    data = tree.climate2,
    aes(x = .data[[clim_x]], y = .data[[clim_y]]),
    probs = c(0.90, 0.75, 0.5)
  ) +

  geom_point(
    data = gom.climate,
    aes(x = .data[[clim_x]], y = .data[[clim_y]]),
    color = "grey20", alpha = 0.3, size = 1
  ) +

  geom_point(
    data = gom.assessed,
    aes(x = .data[[clim_x]], y = .data[[clim_y]], color = performance),
    size = 2, shape = 17, alpha = 0.9
  ) +

  scale_color_manual(
    values = performance_colors,
    drop = FALSE,
    guide = guide_legend(override.aes = list(shape = 17, size = 2))
  ) +

  coord_cartesian(xlim = c(250, 2000), clip = "off") +
  scale_x_continuous(limits = c(250, 2000)) +

  facet_wrap(~ species_full, ncol = 3) +

  theme_minimal(base_size = 10) +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5),
    strip.text = element_blank(),
    strip.background = element_blank(),
    legend.position = "bottom",
    legend.box = "vertical",
    axis.title.x = element_text(size = 8),
    axis.title.y = element_text(size = 8)
  ) +

  labs(
    x = clim_x,
    y = clim_y
  )

hdr.plot

ggsave("./graphics/Species_Climate_Density_w_GoM_tall.png",
       plot = hdr.plot, width = 8.6, height = 24.4, units = "cm")

# -----------------------------
# 5. Species label column
# -----------------------------
species_labels <- paste(names(species_names), "-", species_names)

label_df <- data.frame(
  species_full = species_labels,
  y = seq_along(species_labels)
)

label_plot <- ggplot(label_df, aes(x = 1, y = y, label = species_full)) +
  geom_text(hjust = .2, fontface = "bold", size = 3) +
  scale_y_reverse() +
  coord_cartesian(clip = "off") +
  theme_void() +
  theme(plot.margin = margin(0, 0, 0, 0))

# # -----------------------------
# # 6. Combine with map + HDR
# # -----------------------------
# combined <- (label_plot | hdr.plot| combined_map_plot ) +
#   plot_layout(
#     widths = c(0.15, 0.425, 0.25),
#     guides = "collect"
#   ) &
#   theme(legend.position = "bottom")
# 
# combined

ggsave("./graphics/Species_HDR.png",
       plot = combined, width = 16.8, height = 24.4, units = "cm")
```

Identify climate novelty of GoM trials compared to natural stands

```{r}
clean_numeric <- function(x) {
  x <- as.character(x)
  x <- trimws(x)
  x[x %in% c("", ".", "NA", "NaN", "NULL", "null", "--")] <- NA
  suppressWarnings(as.numeric(x))
}

prep_climate <- function(df, x = "MSP", y = "MCMT") {
  df <- as.data.table(df)
  df[, (x) := clean_numeric(get(x))]
  df[, (y) := clean_numeric(get(y))]
  df
}
library(MASS)
library(FNN)

compute_hdr_membership <- function(tree_sp, gom_sp, 
                                   x = "MSP", y = "MCMT",
                                   prob = 0.95, n = 200) {
  
  # Extract numeric vectors
  refx <- clean_numeric(tree_sp[[x]])
  refy <- clean_numeric(tree_sp[[y]])
  
  kde <- MASS::kde2d(
    x = refx,
    y = refy,
    n = n,
    lims = c(range(refx, na.rm = TRUE), range(refy, na.rm = TRUE))
  )
  
  grid <- expand.grid(
    x = kde$x,
    y = kde$y
  )
  grid$density <- as.numeric(kde$z)
  
  dens_sorted <- sort(grid$density, decreasing = TRUE)
  cum_prob <- cumsum(dens_sorted) / sum(dens_sorted)
  d_thresh <- dens_sorted[which(cum_prob >= prob)[1]]
  
  gom_mat  <- as.matrix(gom_sp[, .(clean_numeric(get(x)), clean_numeric(get(y)))])
  grid_mat <- as.matrix(grid[, c("x", "y")])
  
  nn <- get.knnx(
    data  = grid_mat,
    query = gom_mat,
    k = 1
  )$nn.index
  
  gom_density <- grid$density[nn]
  
  list(
    hdr_flag = gom_density >= d_thresh,
    hdr_density = gom_density,
    d_thresh = d_thresh
  )
}
library(robustbase)

compute_robust_mahal <- function(tree_sp, gom_sp, 
                                 x = "MSP", y = "MCMT") {
  
  tree_mat <- as.matrix(tree_sp[, .(clean_numeric(get(x)), clean_numeric(get(y)))])
  gom_mat  <- as.matrix(gom_sp[,  .(clean_numeric(get(x)), clean_numeric(get(y)))])
  
  mcd <- covMcd(tree_mat)
  center <- mcd$center
  S_inv  <- solve(mcd$cov)
  
  diffs <- sweep(gom_mat, 2, center, "-")
  d2 <- rowSums((diffs %*% S_inv) * diffs)
  sqrt(d2)
}
compute_local_mahal <- function(tree_sp, gom_sp,
                                x = "MSP", y = "MCMT",
                                k_neigh = 20) {
  
  tree_mat <- as.matrix(tree_sp[, .(clean_numeric(get(x)), clean_numeric(get(y)))])
  gom_mat  <- as.matrix(gom_sp[,  .(clean_numeric(get(x)), clean_numeric(get(y)))])
  
  nn_tree <- get.knnx(
    data  = tree_mat,
    query = gom_mat,
    k = 1
  )$nn.index[, 1]
  
  local_mahal <- numeric(nrow(gom_sp))
  
  for (i in seq_len(nrow(gom_sp))) {
    center_idx <- nn_tree[i]
    
    nn_loc <- get.knnx(
      data  = tree_mat,
      query = matrix(tree_mat[center_idx, ], nrow = 1),
      k = min(k_neigh, nrow(tree_mat))
    )$nn.index[1, ]
    
    loc_mat <- tree_mat[nn_loc, , drop = FALSE]
    S <- cov(loc_mat, use = "complete.obs")
    S_inv <- solve(S)
    
    diff <- gom_mat[i, ] - colMeans(loc_mat)
    d2 <- as.numeric(t(diff) %*% S_inv %*% diff)
    local_mahal[i] <- sqrt(d2)
  }
  
  list(
    local_mahal = local_mahal,
    nearest_idx = nn_tree
  )
}
compute_species_climate_novelty <- function(tree_df, gom_df,
                                            x = "MSP", y = "MCMT",
                                            species_col = "species",
                                            hdr_prob = 0.95,
                                            k_neigh = 20) {
  
  tree_df <- prep_climate(tree_df, x, y)
  gom_df  <- prep_climate(gom_df,  x, y)
  
  results <- list()
  
  for (sp in intersect(unique(tree_df[[species_col]]),
                       unique(gom_df[[species_col]]))) {
    
    tree_sp <- tree_df[get(species_col) == sp]
    gom_sp  <- gom_df[get(species_col) == sp]
    
    if (nrow(tree_sp) < 5 || nrow(gom_sp) == 0) next
    
    hdr_res <- compute_hdr_membership(tree_sp, gom_sp, x, y, prob = hdr_prob)
    rob_mahal <- compute_robust_mahal(tree_sp, gom_sp, x, y)
    local_res <- compute_local_mahal(tree_sp, gom_sp, x, y, k_neigh)
    
    nearest_idx <- local_res$nearest_idx
    
    out <- data.table(
      species          = sp,
      gom_PlotNumber   = gom_sp$PlotNumber,
      gom_MSP          = gom_sp[[x]],
      gom_MCMT         = gom_sp[[y]],
      nearest_PlotNumber = tree_sp$PlotNumber[nearest_idx],
      nearest_MSP        = tree_sp[[x]][nearest_idx],
      nearest_MCMT       = tree_sp[[y]][nearest_idx],
      delta_MSP          = gom_sp[[x]] - tree_sp[[x]][nearest_idx],
      delta_MCMT         = gom_sp[[y]] - tree_sp[[y]][nearest_idx],
      robust_mahalanobis = rob_mahal,
      local_mahalanobis  = local_res$local_mahal,
      hdr_inside         = hdr_res$hdr_flag
    )
    
    results[[sp]] <- out
  }
  
  rbindlist(results, use.names = TRUE, fill = TRUE)
}

```

climate novelty of GoM trials, plot by species with assessed trials highlighted
```{r}
novelty_table <- compute_species_climate_novelty(
  tree_df = tree.climate2,
  gom_df  = gom.climate,
  x = "DD5_an",
  y = "MCMT",
  species_col = "species",
  hdr_prob = 0.95,
  k_neigh = 20
)
fwrite(novelty_table, "./clean_data/GoM_trials_climate_novelty.csv")
dist_stats <- novelty_table %>%
  group_by(species) %>%
  summarise(
    n1 = n(),
    min = min(local_mahalanobis, na.rm = TRUE),
    q1 = quantile(local_mahalanobis, 0.25, na.rm = TRUE),
    median = median(local_mahalanobis, na.rm = TRUE),
    mean = mean(local_mahalanobis, na.rm = TRUE),
    q3 = quantile(local_mahalanobis, 0.75, na.rm = TRUE),
    max = max(local_mahalanobis, na.rm = TRUE)
  )
dist_stats_gt <- dist_stats %>%
  gt(rowname_col = "species") %>%
  tab_header(
    title = "Distribution of Local Mahalanobis Distance by Species"
  ) %>%
  fmt_number(
    columns = c(min, q1, median, mean, q3, max),
    decimals = 2
  ) %>%
  cols_label(
    n1 = "N",
    min = "Min",
    q1 = "Q1",
    median = "Median",
    mean = "Mean",
    q3 = "Q3",
    max = "Max"
  ) %>%
  tab_options(
    table.font.size = px(12),
    data_row.padding = px(3)
  )

dist_stats_gt

dist_stats
novelty_classes <- novelty_table %>%
  mutate(
    novelty_class = case_when(
      local_mahalanobis < 1 ~ "in_range",
      local_mahalanobis >= 1 & local_mahalanobis <= 3 ~ "peripheral",
      local_mahalanobis > 3 ~ "novel"
    )
  ) %>%
  group_by(species, novelty_class) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(species) %>%
  mutate(prop = n / sum(n))

novelty_classes

combined_summary <- dist_stats %>%
  left_join(
    novelty_classes %>%
      tidyr::pivot_wider(
        names_from = novelty_class,
        values_from = prop,
        values_fill = 0
      ),
    by = "species"
  )

combined_summary
combined_gt <- combined_summary %>%
  gt(rowname_col = "species") %>%
  tab_header(
    title = "Climate Novelty Summary by Species"
  ) %>%
  fmt_number(
    columns = c(min, q1, median, mean, q3, max),
    decimals = 2
  ) %>%
  fmt_percent(
    columns = c(in_range, peripheral, novel),
    decimals = 1
  ) %>%
  cols_label(
    n = "N",
    min = "Min",
    q1 = "Q1",
    median = "Median",
    mean = "Mean",
    q3 = "Q3",
    max = "Max",
    in_range = "< 1 (In‑range)",
    peripheral = "1–3 (Peripheral)",
    novel = "> 3 (Novel)"
  ) %>%
  tab_options(
    table.font.size = px(12),
    data_row.padding = px(3)
  )

combined_gt

```

climate novelty of GoM trials

```{r}
# -----------------------------
# 1. Define climate variables
# -----------------------------
clim_x <- "DD5_an"
clim_y <- "MCMT"

# -----------------------------
# 2. Build assessed GoM table
# -----------------------------
gom.assessed2 <- assessed_gom_trials %>%
  filter(PERIOD == "1961_1990",
         species %in% c("Cw", "Fd", "Lw", "Pw", "Py")) %>%
  mutate(PlotNumber = paste0(PlotNumber, "_", species, "_", replicate)) %>%
  dplyr::select(
    PlotNumber = id,
    species,
    mean_vig,
    survival,
    performance,
    !!clim_x := .data[[clim_x]],
    !!clim_y := .data[[clim_y]],
    nearest_natural_km
  ) %>%
  mutate(
    performance = as.character(performance),
    PlotNumber = as.character(PlotNumber)
  )

# -----------------------------
# 3. Compute novelty
# -----------------------------
novelty_table.ass <- compute_species_climate_novelty(
  tree_df = tree.climate2,
  gom_df  = gom.assessed2,
  x = clim_x,
  y = clim_y,
  species_col = "species",
  hdr_prob = 0.95,
  k_neigh = 20
) %>%
  data.frame() %>%
  mutate(PlotNumber = gom_PlotNumber)

novelty_table.ass <- left_join(
  gom.assessed2,
  novelty_table.ass,
  by = c("PlotNumber", "species")
) %>% distinct()

# -----------------------------
# 4. Distance plot
# -----------------------------
distance_plot <- ggplot(summary_data,
                        aes(x = species, y = nearest_natural_km, fill = species)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.2, alpha = 0.5, color = "black", size = .5) +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "none",
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    plot.title = element_text(face = "bold")
  ) +
  labs(
    x = "Species",
    y = "Nearest Natural Stand (km)",
    color = "Performance"
  )

ggsave("./graphics/GoM_trial_distances_to_natural_plots.png",
       plot = distance_plot, width = 8.6, height = 8, units = "cm")

# -----------------------------
# 5. Climate novelty plot
# -----------------------------
climate.plot <- ggplot() +
  geom_boxplot(
    data = novelty_table,
    aes(x = species, y = local_mahalanobis, fill = species),
    outlier.shape = NA,
    outlier.alpha = 0.2,
    width = 0.6
  ) +
  geom_jitter(
    data = novelty_table,
    aes(x = species, y = local_mahalanobis),
    width = 0.2,
    alpha = 0.5,
    size = .5,
    color = "black"
  ) +
  geom_jitter(
    data = novelty_table.ass,
    aes(x = species, y = local_mahalanobis, color = performance),
    width = 0.15,
    alpha = 0.9,
    size = 1,
    shape = 17
  ) +
  scale_color_manual(
    values = performance_colors,
    drop = FALSE,
    guide = guide_legend(override.aes = list(shape = 17, size = 2))
  ) +
  geom_hline(yintercept = 1, linetype = "dashed",
             color = "forestgreen", linewidth = 1) +
  geom_hline(yintercept = 3, linetype = "dashed",
             color = "firebrick", linewidth = 1) +
  scale_y_continuous(
    breaks = seq(0, max(novelty_table$local_mahalanobis, na.rm = TRUE), by = 1)
  ) +
  theme_minimal(base_size = 8) +
  theme(
    legend.position = "none",
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    plot.title = element_text(face = "bold")
  ) +
  labs(
    x = "Species",
    y = "SD Units (Local Mahalanobis)",
    title = paste0("Local Mahalanobis Novelty (", clim_x, " vs ", clim_y, ")"),
    color = "Performance"
  )

ggsave("./graphics/GoM_assessed_trials_local_mahalanobis_by_species.png",
       plot = climate.plot, width = 8.6, height = 8, units = "cm")

# -----------------------------
# 6. HDR panel grid
# -----------------------------
hdr.plot <- plot_grid(
  hdr_panels$Cw,
  hdr_panels$Fd,
  hdr_panels$Lw,
  hdr_panels$Pw,
  hdr_panels$Py,
  climate.plot,
  ncol = 3,
  nrow = 2,
  labels = c("A", "B", "C", "D", "E", "F"),
  label_size = 12,
  label_fontface = "bold",
  label_x = 0.02,
  label_y = 0.98,
  hjust = 0,
  vjust = 1,
  align = "hv",
  axis = "tblr"
)

ggsave("./graphics/HDR_GoM_mahalanobis_by_species.png",
       plot = hdr.plot, width = 24.4, height = 16, units = "cm")
```
## Performance by novelty class plot
```{r}
novelty_table.ass <- novelty_table.ass %>% filter(!is.na(performance)) %>% 
  mutate(
    novelty_class = case_when(
      local_mahalanobis < 1 ~ "In-range",
      local_mahalanobis < 3 ~ "Peripheral",
      TRUE ~ "Novel"
    ),
    novelty_class = factor(novelty_class,
                           levels = c("In-range", "Peripheral", "Novel")),
    performance = factor(performance,
                         levels = c("Fail", "Poor", "Fair", "Good", "Excellent"))
  )
if ("sf" %in% class(novelty_table.ass)) {
  novelty_table.ass <- sf::st_drop_geometry(novelty_table.ass)
}

perf_summary_species <- novelty_table.ass %>%
  group_by(species, novelty_class, performance) %>%
  summarise(n = n(), .groups = "drop_last") %>%
  mutate(
    total = sum(n),
    proportion = n / total
  ) %>%
  ungroup() %>%
  dplyr::select(species, novelty_class, performance, n, proportion)

perf_summary_species_gt <- perf_summary_species %>%
  gt(
    rowname_col = "performance",
    groupname_col = "novelty_class"
  ) %>%
  tab_spanner(
    label = "Species",
    columns = species
  ) %>%
  fmt_percent(columns = proportion, decimals = 1) %>%
  cols_label(
    species = "Species",
    n = "Count",
    proportion = "Proportion"
  ) %>%
  tab_header(
    title = "Performance Ratios by Novelty Class and Species"
  )

perf_summary_species_gt


perf_novelty_plot <- novelty_table.ass %>%
  ggplot(aes(x = novelty_class, fill = performance)) +
  geom_bar(position = "fill", color = "black", linewidth = 0.2) +
  scale_fill_manual(values = performance_colors, drop = FALSE) +
  facet_wrap(~ species, ncol = 3) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_minimal(base_size = 8) +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold"),
    legend.position = "bottom"
  ) +
  labs(
    x = "Novelty Class",
    y = "Proportion of Assessed Trials",
    fill = "Performance"#,
    #title = paste0("Performance Ratings by Novelty Class (", clim_x, " vs ", clim_y, ")")
  )

perf_novelty_plot

ggsave("./graphics/GoM_performance_by_novelty_class.png",
       plot = perf_novelty_plot, width = 8.6, height = 8.6, units = "cm")


```


```{r}
gom.assessed2 <- assessed_gom_trials %>% filter(PERIOD == "1961_1990", species %in% c("Cw", "Fd", "Lw", "Pw", "Py")) %>% mutate(PlotNumber = paste0(PlotNumber, "_", species, "_", replicate)) %>%  dplyr::select(PlotNumber =id, species, mean_vig, survival, performance, DD5_an, MCMT, nearest_natural_km) %>% 
  mutate(performance = as.character(performance),
         PlotNumber = as.character(PlotNumber))
# assessed_distances <- assessed_gom_trials_with_distances %>% 
#   st_drop_geometry() %>%
#   dplyr::select(PlotNumber, species, nearest_natural_km) 
novelty_table.ass <- compute_species_climate_novelty(
  tree_df = tree.climate2,
  gom_df  = gom.assessed2,
  x = "DD5_an",
  y = "MCMT",
  species_col = "species",
  hdr_prob = 0.95, 
  k_neigh = 20
)
gom.assessed2 <- gom.assessed2  %>% data.frame #%>% mutate(PlotNumber = as.factor(PlotNumber))
novelty_table.ass <- novelty_table.ass  %>% data.frame %>% mutate(PlotNumber = gom_PlotNumber)
novelty_table.ass <- left_join(gom.assessed2, novelty_table.ass, by = c("PlotNumber", "species")) %>% distinct

distance_plot <- ggplot(summary_data, aes(x = species, y = nearest_natural_km, fill = species)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.2, alpha = 0.5, color = "black", size = .5) +
  #scale_y_continuous(name = "Nearest Natural Plot (km)") +
  #scale_x_discrete(name = "Species") +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "none",
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    plot.title = element_text(face = "bold")
  ) +
    labs(
    x = "Species",
    y = "Nearest Natural Stand (km)",
    #title = "Local Mahalanobis Novelty by Species",
    color = "Performance" )
#+
distance_plot #labs(title = "Distance to Nearest Natural Stand")
ggsave("./graphics/GoM_trial_distances_to_natural_plots.png", plot = distance_plot, width = 8.6, height = 8, units = "cm")

climate.plot <- ggplot() +
  
  # Boxplot for all trials
  geom_boxplot(
    data = novelty_table,
    aes(x = species, y = local_mahalanobis, fill = species),
    outlier.shape = NA,
    outlier.alpha = 0.2,
    width = 0.6
  ) +
  
  # Jitter for all trials (grey)
  geom_jitter(
    data = novelty_table,
    aes(x = species, y = local_mahalanobis),
    width = 0.2,
    alpha = 0.5,
    size = .5,
    color = "black"
  ) +
  
  # Jitter for assessed trials (colored by performance)
  geom_jitter(
    data = novelty_table.ass,
    aes(x = species, y = local_mahalanobis, color = performance),
    width = 0.15,
    alpha = 0.9,
    size = 1,
    shape = 17
  ) +
  scale_color_manual(
    values = performance_colors,
    drop = FALSE,
    guide = guide_legend(override.aes = list(shape = 17, size = 2))
  ) +
  
  # Horizontal cutoffs
  geom_hline(yintercept = 1, linetype = "dashed", color = "forestgreen", linewidth = 1) +
  geom_hline(yintercept = 3, linetype = "dashed", color = "firebrick", linewidth = 1) +
  
  # Labels for cutoffs (bold)
# annotate("label", x = 1.32, y = 0,
#          label = "in range",
#          vjust = 0, hjust = 0, size = 3,
#          color = "forestgreen", fontface = "bold",
#          fill = "white", alpha = 0.8, label.size = 0) +
# annotate("label", x = 1.32, y = 2,
#          label = "peripheral",
#          vjust = 0.5, hjust = 0, size = 3,
#          color = "goldenrod", fontface = "bold",
#          fill = "white", alpha = 0.8, label.size = 0) +
# annotate("label", x = 1.32, y = 3.5,
#          label = "novel",
#          vjust = 0, hjust = 0, size = 3,
#          color = "firebrick", fontface = "bold",
#          fill = "white", alpha = 0.8, label.size = 0) +
  
  # Y-axis ticks every 1
  scale_y_continuous(
    breaks = seq(0, max(novelty_table$local_mahalanobis, na.rm = TRUE), by = 1)
  ) +
  
  # Theme cleanup
  theme_minimal(base_size = 8) +
  theme(
    legend.position = "none",
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    plot.title = element_text(face = "bold")
  ) +
  
  labs(
    x = "Species",
    y = "SD Units (Local Mahalanobis)",
    title = "Local Mahalanobis Novelty by Species",
    color = "Performance"
  )
climate.plot
ggsave("./graphics/GoM_assessed_trials_local_mahalanobis_by_species.png", plot = climate.plot, width = 8.6, height = 8, units = "cm")


hdr.plot <- plot_grid(
  hdr_panels$Cw,   # A
  hdr_panels$Fd,   # B
  hdr_panels$Lw,   # C
  hdr_panels$Pw,   # D
  hdr_panels$Py,   # E
  climate.plot,   # F
  ncol = 3,
  nrow = 2,
  labels = c("A", "B", "C", "D", "E", "F"),
  label_size = 12,
  label_fontface = "bold",
  label_x = 0.02,   # inside border
  label_y = 0.98,   # inside border
  hjust = 0,
  vjust = 1,
  align = "hv",
  axis = "tblr"
)

hdr.plot
ggsave("./graphics/HDR_GoM_mahalanobis_by_species.png", plot = hdr.plot, width = 24.4, height = 16, units = "cm")
```


```{r}
gom.assessed2 <- gom.assessed2  %>% data.frame #%>% mutate(PlotNumber = as.factor(PlotNumber))
novelty_table.ass <- novelty_table.ass  %>% data.frame %>% mutate(PlotNumber = gom_PlotNumber)
xx <- left_join(gom.assessed2, novelty_table.ass, by = c("PlotNumber", "species")) %>% distinct

library(dplyr)

performance_by_hdr <- xx %>%
  group_by(species, hdr_inside) %>%
  summarise(
    n = n(),
    mean_performance = mean(performance, na.rm = TRUE),
   mean_dist   = mean(nearest_natural_km, na.rm = TRUE),
    mean_survival    = mean(survival, na.rm = TRUE),
    mean_vig         = mean(mean_vig, na.rm = TRUE),
    .groups = "drop"
  )
library(ggplot2)
#assessed <- fread( "./clean_data/GoM_assessed_trials_2023_24.csv")
# Prepare data with counts and percentages
performance_summary <- xx %>%
  filter(species %in% c("Cw", "Fd", "Lw", "Pw", "Py")) %>%
  drop_na() %>%
  count(species, hdr_inside, performance) %>%   # gives n
  group_by(species, hdr_inside) %>%             # percent within HDR class
  mutate(percent = n / sum(n) * 100) %>%
  ungroup() %>%
  mutate(performance = factor(
    performance,
    levels = c("Excellent", "Good", "Fair", "Poor", "Fail"),
    ordered = TRUE
  ))


ggplot(performance_summary, aes(x = species, y = n, fill = performance)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(
    aes(label = paste0(round(percent), "%")),
    position = position_stack(vjust = 0.5),
    size = 3,
    color = "black"
  ) +
  theme_minimal() +
  labs(
    x = "Species",
    y = "Count of Trials",
    fill = "Performance"
  ) +
  scale_fill_manual(values = c(
    "Excellent" = "purple",
    "Good"      = "green",
    "Fair"      = "yellow",
    "Poor"      = "orange",
    "Fail"      = "red"
  )) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(size = 10, face = "bold"),
    strip.text  = element_text(size = 10, face = "bold")
  ) +
  facet_wrap(~ hdr_inside)


ggsave("./graphics/GoM_assessed_trials_performance_2023_24.png", width = 7.3725, height = 4)
```
library(ggplot2)

ggplot(xx, aes(x = delta_MCMT, y = local_mahalanobis)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  facet_wrap(~ species, scales = "free") +
  labs(
    x = "MCMT distance (GoM – nearest treeplot)",
    y = "Robust Mahalanobis",
    title = "RM vs delta_MCMT Climate Distance by Species"
  ) +
  theme_bw()

ggplot(xx, aes(x = local_mahalanobis, y = mean_vig)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  geom_hline(yintercept = 2.5, linetype = "dashed", color = "darkgreen") +
  facet_wrap(~ species, scales = "free") +
  labs(
    x = "Robust Mahalanobis",
    y = "Mean Vigor",
    title = "vigour vs delta_MCMT Climate Distance by Species"
  ) +
  theme_bw()
ggplot(xx, aes(x = local_mahalanobis, y = mean_vig)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  geom_hline(yintercept = 2.5, linetype = "dashed", color = "darkgreen") +
  facet_wrap(~ species, scales = "free") +
  labs(
    x = "Robust Mahalanobis",
    y = "Mean Vigor",
    title = "RM vs delta_MCMT Climate Distance by Species"
  ) +
  theme_bw()

```


####OLD CODE###################


```{r }
# library(sf)
# library(ggplot2)
# library(dplyr)
# library(data.table)
# 
# tree.climate2 <- fread("./clean_data/Tree_Plot_Climate.csv")
# tree.sf <- st_as_sf(tree.climate2, coords = c("Longitude", "Latitude"), crs = 4326) %>%
#   left_join(tree.codes, by = c("Species" = "Code"))
# 
# gom.sf <- st_as_sf(gom.climate, coords = c("longitude", "latitude"), crs = 4326) %>%
#   filter(species %in% c("Cw", "Fd", "Lw", "Pw", "Py"))
# 
# provmap <- st_read("D:/OneDrive - Government of BC/CommonTables/BC_AB_US_Shp/BC_Province_Outline_Dissolved.gpkg")
# 
# gom.assessed <- fread("./clean_data/GoM_assessed_trials_2023_24.csv")
# gom.assessed.sf <- st_as_sf(gom.assessed, coords = c("Block_long", "Block_lat"), crs = 4326)
# 
# # Define performance levels and colors
# performance_levels <- c("Fail", "Poor", "Fair", "Good", "Excellent")
# performance_colors <- c(
#   "Fail" = "red",
#   "Poor" = "orange",
#   "Fair" = "yellow",
#   "Good" = "green",
#   "Excellent" = "purple"
# )
#  #sp = "Cw"
# for (sp in unique(gom.sf$species)) {
#   gom_data <- gom.sf %>% filter(species == sp)
#   tree_data <- tree.sf %>% filter(species == sp)
#   assessed_data <- gom.assessed.sf %>%
#     filter(species == sp) %>%
#     mutate(performance = factor(performance, levels = performance_levels))
# 
#   # Apply jitter to assessed_data coordinates
#   jittered_coords <- st_coordinates(assessed_data) +
#     matrix(runif(nrow(assessed_data) * 2, -0.03, 0.03), ncol = 2)
#   assessed_data_jittered <- st_as_sf(
#     cbind(assessed_data, jittered_coords),
#     coords = c("X", "Y"),
#     crs = 4326
#   )
# 
#   p <- ggplot() +
#     geom_sf(data = provmap, fill = "grey90", color = "black") +
#     geom_sf(data = tree_data, color = "grey50", size = 1, alpha = 0.8, shape = 1) +
#     geom_sf(data = gom_data, color = "grey20", size = 1, alpha = 0.8, shape = 15) +
#     geom_sf(data = assessed_data_jittered, aes(color = performance), size = 3, shape = 17, alpha = 0.9) +
#     scale_color_manual(values = performance_colors, drop = FALSE) +
#     coord_sf(xlim = st_bbox(provmap)[c("xmin", "xmax")],
#              ylim = st_bbox(provmap)[c("ymin", "ymax")],
#              expand = FALSE) +
#     theme_minimal()+
#     labs(
#       title = paste("Tree Plot Locations for", sp),
#       subtitle = "Species Distribution with Climate Data and Assessed Performance",
#       color = "Performance"
#     )
#   print(p)
#   ggsave(filename = paste0("./graphics/map_", sp, ".png"), plot = p, width = 8, height = 6)
# }

# 
# tree_data <- tree.sf %>% filter(species == "Fd")
# ggplot() +
#     geom_sf(data = provmap, fill = "grey90", color = "black") +
#     geom_sf(data = tree_data, color = "grey50", size = 1, alpha = 0.8, shape = 1) +
#     #geom_sf(data = gom_data, color = "grey20", size = 2, alpha = 0.8, shape = 15) +
#     #geom_sf(data = assessed_data_jittered, aes(color = performance), size = 3, shape = 17, alpha = 0.9) +
#     #scale_color_manual(values = performance_colors, drop = FALSE) +
#     coord_sf(xlim = st_bbox(provmap)[c("xmin", "xmax")],
#              ylim = st_bbox(provmap)[c("ymin", "ymax")],
#              expand = FALSE) +
#     theme_minimal()+
#     labs(
#       title = paste("Tree Plot Locations for", sp),
#       subtitle = "Species Distribution with Climate Data and Assessed Performance",
#       color = "Performance"
#     )
```

```{r}
# library(sf)
# library(dplyr)
# library(data.table)
# library(geosphere)  # For geodesic distance calculation
# 
# # Load and prepare spatial data (already done in your script)
# tree.climate2 <- fread("./clean_data/Tree_Plot_Climate.csv")
# tree.sf <- st_as_sf(tree.climate2, coords = c("Longitude", "Latitude"), crs = 4326) %>%
#   left_join(tree.codes, by = c("Species" = "Code"))
# 
# gom.sf <- st_as_sf(gom.climate, coords = c("longitude", "latitude"), crs = 4326) %>%
#   filter(species %in% c("Cw", "Fd", "Lw", "Pw", "Py"))
# 
# gom.assessed <- fread("./clean_data/GoM_assessed_trials_2023_24.csv")
# gom.assessed.sf <- st_as_sf(gom.assessed, coords = c("Block_long", "Block_lat"), crs = 4326)
# 
# # Function to calculate nearest distance (in km) from each trial to natural stand
# calculate_nearest_distance <- function(trial_sf, natural_sf) {
#   trial_coords <- st_coordinates(trial_sf)
#   natural_coords <- st_coordinates(natural_sf)
# 
#   nearest_distances <- sapply(1:nrow(trial_coords), function(i) {
#     trial_point <- trial_coords[i, ]
#     dists <- distHaversine(trial_point, natural_coords) / 1000  # Convert to km
#     min(dists)
#   })
# 
#   trial_sf$nearest_natural_km <- nearest_distances
#   return(trial_sf)
# }
# 
# # Loop over species and calculate distances
# distance_results <- list()
# 
# for (sp in unique(gom.sf$species)) {
#   tree_data <- tree.sf %>% filter(species == sp)
#   trial_data <- gom.sf %>% filter(species == sp)
# 
#   if (nrow(tree_data) > 0 && nrow(trial_data) > 0) {
#     trial_with_dist <- calculate_nearest_distance(trial_data, tree_data)
#     distance_results[[sp]] <- trial_with_dist
#   }
# }
# 
# # Combine all species results into one sf object
# gom_trials_with_distances <- do.call(rbind, distance_results)
# 
# # Optional: Save to CSV or shapefile
# # st_write(gom_trials_with_distances, "./output/gom_trials_with_distances.gpkg")
# # fwrite(st_drop_geometry(gom_trials_with_distances), "./output/gom_trial_distances.csv")
# 
# library(ggplot2)
# library(dplyr)
# 
# # Ensure 'species' and 'nearest_natural_km' are present
# summary_data <- gom_trials_with_distances %>%
#   st_drop_geometry() %>%
#   filter(!is.na(nearest_natural_km)) %>%
#   mutate(species = factor(species))  # Optional: reorder by median distance
# 
# # Plot: Boxplot with jittered points
# ggplot(summary_data, aes(x = species, y = nearest_natural_km, fill = species)) +
#   geom_boxplot(outlier.shape = NA, alpha = 0.6) +
#   geom_jitter(width = 0.2, alpha = 0.5, color = "black", size = 1.5) +
#   scale_y_continuous(name = "Distance to Nearest Natural Plot (km)") +
#   scale_x_discrete(name = "Species") +
#   theme_minimal(base_size = 14) +
#   theme(legend.position = "none") +
#   labs(
#     title = "Distance from GoM Trials to Nearest Natural Plot",
#     subtitle = "By Species",
#     caption = "Distances calculated using geodesic (Haversine) method"
#   )
# 
# ggsave("./graphics/GoM_trial_distances_to_natural_plots.png", width = 8.6, height = 6, units = "cm")
```


```{r}
# library(dplyr)
# library(gt)
# 
# # Drop geometry and filter valid distances
# summary_data <- gom_trials_with_distances %>%
#   st_drop_geometry() %>%
#   filter(!is.na(nearest_natural_km)) %>%
#   group_by(species) %>%
#   summarise(
#     Min_Distance_km = round(min(nearest_natural_km), 1),
#     Median_Distance_km = round(median(nearest_natural_km), 1),
#     Max_Distance_km = round(max(nearest_natural_km), 1),
#     .groups = "drop"
#   )
# 
# # Optional: Replace species codes with common names
# species_lookup <- c(
#   "Cw" = "Western redcedar",
#   "Fd" = "Douglas-fir",
#   "Lw" = "Western larch",
#   "Pw" = "Western white pine",
#   "Py" = "Ponderosa pine"
# )
# 
# summary_data <- summary_data %>%
#   mutate(Species = species_lookup[species]) %>%
#   select(Species, Min_Distance_km, Median_Distance_km, Max_Distance_km)
# 
# # Create gt table
# gt_table <- summary_data %>%
#   gt() %>%
#   tab_header(
#     title = "Distance from GoM Trials to Nearest Natural Plot",
#     subtitle = "Minimum, Median, and Maximum by Species"
#   ) %>%
#   fmt_number(columns = everything(), decimals = 1) %>%
#   cols_label(
#     Min_Distance_km = "Min (km)",
#     Median_Distance_km = "Median (km)",
#     Max_Distance_km = "Max (km)"
#   ) %>%
#   tab_options(
#     table.font.size = "small",
#     data_row.padding = px(4)
#   )
# 
# # Print table
# gt_table
# # Save as PNG (requires webshot2 or ragg)
# gtsave(gt_table, filename = "./graphics/gom_distance_summary.png")

```

table to check for errors
```{r}
# library(sf)
# library(dplyr)
# library(data.table)
# library(geosphere)
#
# # Replace with actual column names
# gom_plot_col <- "PlotNumber"     # GoM trial plot identifier
# tree_plot_col <- "PlotNumber"    # Natural tree plot identifier
#
# # Function to find nearest tree plot for each GoM trial
# find_nearest_plot <- function(trial_sf, tree_sf, trial_id_col, tree_id_col) {
#   trial_coords <- st_coordinates(trial_sf)
#   tree_coords <- st_coordinates(tree_sf)
#
#   nearest_info <- lapply(1:nrow(trial_coords), function(i) {
#     trial_point <- trial_coords[i, ]
#     dists <- distHaversine(trial_point, tree_coords)
#     nearest_index <- which.min(dists)
#     tibble(
#       GoM_PlotNumber = trial_sf[[trial_id_col]][i],
#       Species = trial_sf$species[i],
#       Nearest_Tree_PlotNumber = tree_sf[[tree_id_col]][nearest_index],
#       Distance_km = round(dists[nearest_index] / 1000, 2)
#     )
#   })
#
#   bind_rows(nearest_info)
# }
#
# # Loop over species
# nearest_table <- list()
#
# for (sp in unique(gom.sf$species)) {
#   trial_data <- gom.sf %>% filter(species == sp)
#   tree_data <- tree.sf %>% filter(species == sp)
#
#   if (nrow(trial_data) > 0 && nrow(tree_data) > 0) {
#     nearest_table[[sp]] <- find_nearest_plot(trial_data, tree_data, gom_plot_col, tree_plot_col)
#   }
# }
#
# # Combine all species into one table
# gom_nearest_plot_table <- bind_rows(nearest_table)
#
# # View or export
# print(gom_nearest_plot_table)
# fwrite(gom_nearest_plot_table, "./output/gom_nearest_plot_table.csv")
```

```{r}
# library(ggplot2)
# 
# # Create a data frame of shape codes
# shape_df <- data.frame(
#   shape = 0:25,
#   x = rep(1:6, length.out = 26),
#   y = rep(5:1, each = 6, length.out = 26)
# )
# 
# # Plot the shape codes
# ggplot(shape_df, aes(x = x, y = y)) +
#   geom_point(aes(shape = factor(shape)), size = 5, fill = "steelblue") +
#   geom_text(aes(label = shape), vjust = -1.5, size = 4) +
#   scale_shape_manual(values = shape_df$shape) +
#   theme_void() +
#   ggtitle("ggplot2 Shape Codes (0–25)")
# 
# library(colorspace)
# 
# # Choose a qualitative palette
# qualitative_hcl(5, palette = "Set 2")
# 
# # Visualize multiple palettes
# hcl_palettes(plot = TRUE)

# Interactive palette picker
#choose_palette()  # Opens a GUI if you're in RStudio

```


```{r}
# Load and filter data
# library(data.table)
# # bad.locations <- fread("./raw_data/Bad_Plot_Locations.csv") %>% pull(PlotNumber)
# # tree.climate2 <- fread("./clean_data/Tree_Plot_Climate.csv") %>% filter(!PlotNumber %in% bad.locations)
# # fwrite(tree.climate2, "./clean_data/Tree_Plot_Climate.csv")
# tree.climate2 <- fread("./clean_data/Tree_Plot_Climate.csv") %>% 
#   filter(species %in% c("Fd", "Lw", "Py", "Pw", "Cw")) %>% #CoastInterior == "Interior", 
#   distinct()#, "Bg""At", "Hw", 
# 
# gom.climate <- fread("./clean_data/GoM_trial_climate.csv") %>%
#   filter(species %in% c("Fd", "Lw", "Py", "Pw", "Cw"))#,  "Bg""At", "Hw",
# 
# WNA.climate <- fread("D:/OneDrive - Government of BC/CommonTables/BGC_climate/WNA_BGCv13_10km.csv") %>%
#   filter(!is.na(MAT), !is.na(MAP), !is.na(MSP), !is.na(MWMT), !is.na(CMD_an), !is.na(DD5_an))
# 
# gom.assessed <- fread("./clean_data/GoM_assessed_climate.csv") %>%
#   filter(species %in% c("Cw","Fd","Lw",  "Pw","Py" )) %>% arrange(species) %>% 
#   mutate(performance = factor(performance, levels = c("Fail", "Poor", "Fair", "Good", "Excellent"))) %>% distinct %>% drop_na(performance)
# 
# # Define performance levels and colors
# performance_levels <- c("Fail", "Poor", "Fair", "Good", "Excellent")
# performance_colors <- c(
#   "Fail" = "red",
#   "Poor" = "orange",
#   "Fair" = "yellow",
#   "Good" = "green",
#   "Excellent" = "purple")
# species_names <- c(
#   "Cw" = "Western redcedar",
#   "Fd" = "Douglas-fir",
#   "Lw" = "Western larch",
#   "Pw" = "Western white pine",  
#   "Py" = "Ponderosa pine"
# )
# tree.climate2[, species_full := species_names[species]]
# gom.assessed[, species_full := species_names[species]]
# gom.climate[, species_full := species_names[species]]
# 
# ggplot(tree.climate2, aes(x = MSP, y = MCMT)) +
#   geom_hdr(data = tree.climate2, aes(x = MSP, y = MCMT), probs = c(0.90, 0.75, 0.5)) +
#   geom_point(data = gom.climate, aes(x = MSP, y = MCMT), color = "grey20", alpha = 0.3, size = 1) +
#   geom_point(
#     data = gom.assessed,
#     aes(x = MSP, y = MCMT, color = performance),
#     size = 2, shape = 17, alpha = 0.9
#   ) +
#   scale_color_manual(
#     values = performance_colors,
#     drop = FALSE,
#     guide = guide_legend(override.aes = list(shape = 17, size = 3))
#   ) +
#   coord_cartesian(clip = "off") +
#   facet_wrap(~ species_full, ncol = 3) +
#   theme_minimal(base_size = 12) +
#   theme(
#     panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5),
#     strip.text = element_text(face = "bold", size = 12)
#   ) +
#   labs(
#     #title = "Species-Specific Density Across Climate Space",
#     #subtitle = "Western North America",
#     x = "MSP", y = "MCMT"
#   )
# 
# ggsave("./graphics/Species_Climate_Density_w_GoM.png", width = 24.4, height = 16 , units = "cm")
#  xx <- gom.climate %>% filter(species == "Cw", MSP > 550 & MCMT < -9) %>%  dplyr::select(PERIOD, PlotNumber, species, longitude, latitude, MSP, MCMT)
```


```{r}
# library(FNN)
# library(dplyr)
# 
# compute_climate_nearest_treeplot <- function(tree_df, gom_df, x = "MSP", y = "MCMT") {
#   
# # Force numeric
# tree_df[[x]] <- as.numeric(tree_df[[x]])
# tree_df[[y]] <- as.numeric(tree_df[[y]])
# gom_df[[x]]  <- as.numeric(gom_df[[x]])
# gom_df[[y]]  <- as.numeric(gom_df[[y]])
# 
# # Correct data.table column extraction
# cols <- c(x, y)
# tree_mat <- as.matrix(tree_df[, ..cols])
# gom_mat  <- as.matrix(gom_df[,  ..cols])
#   # 1 nearest neighbour
#   nn <- get.knnx(
#     data  = tree_mat,
#     query = gom_mat,
#     k = 1
#   )
#   
#   idx <- nn$nn.index[, 1]
#   
#   # Extract nearest treeplot climate
# nearest <- tree_df[idx, cols, with = FALSE]
# 
#   names(nearest) <- c("nearest_MSP", "nearest_MCMT")
#   
#   # Compute deltas
#   gom_df$delta_MSP  <- gom_df[[x]] - nearest$nearest_MSP
#   gom_df$delta_MCMT <- gom_df[[y]] - nearest$nearest_MCMT
#   
#   # Euclidean climate distance
#   gom_df$climate_distance <- sqrt(gom_df$delta_MSP^2 + gom_df$delta_MCMT^2)
#   
#   # Add nearest treeplot ID if available
#   if ("PlotNumber" %in% names(tree_df)) {
#     gom_df$nearest_treeplot <- tree_df$PlotNumber[idx]
#   }
#   
#   return(gom_df)
# }
# 
# library(data.table)
# 
# compute_species_mahalanobis <- function(tree_df, gom_df,
#                                                   x = "MSP", y = "MCMT",
#                                                   species_col = "species") {
#   
#   tree_df <- as.data.table(tree_df)
#   gom_df  <- as.data.table(gom_df)
#   
#   # Force numeric
#   tree_df[, (x) := as.numeric(get(x))]
#   tree_df[, (y) := as.numeric(get(y))]
#   gom_df[,  (x) := as.numeric(get(x))]
#   gom_df[,  (y) := as.numeric(get(y))]
#   
#   results <- list()
#   
#   for (sp in unique(gom_df[[species_col]])) {
#     
#     gom_sp  <- gom_df[get(species_col) == sp]
#     tree_sp <- tree_df[get(species_col) == sp]
#     
#     if (nrow(tree_sp) < 3) next   # need at least 3 points for covariance
#     
#     # Covariance matrix
#     S <- cov(tree_sp[, .(get(x), get(y))], use = "complete.obs")
#     S_inv <- solve(S)
#     
#     # Matrices
#     tree_mat <- as.matrix(tree_sp[, .(get(x), get(y))])
#     gom_mat  <- as.matrix(gom_sp[,  .(get(x), get(y))])
#     
#     # Compute Mahalanobis distances for treeplots (to define envelope)
#     tree_diffs <- sweep(tree_mat, 2, colMeans(tree_mat), "-")
#     tree_d2 <- rowSums((tree_diffs %*% S_inv) * tree_diffs)
#     tree_mahal <- sqrt(tree_d2)
#     
#     # 95% envelope threshold
#     d95 <- quantile(tree_mahal, 0.95, na.rm = TRUE)
#     
#     # Compute Mahalanobis distances for GoM
#     nearest_idx <- integer(nrow(gom_mat))
#     mahal_dist  <- numeric(nrow(gom_mat))
#     
#     for (i in seq_len(nrow(gom_mat))) {
#       diffs <- sweep(tree_mat, 2, gom_mat[i, ], "-")
#       d2 <- rowSums((diffs %*% S_inv) * diffs)
#       nearest_idx[i] <- which.min(d2)
#       mahal_dist[i]  <- sqrt(min(d2))
#     }
#     
#     out <- data.table(
#       species = sp,
#       gom_PlotNumber = gom_sp$PlotNumber,
#       gom_MSP  = gom_sp[[x]],
#       gom_MCMT = gom_sp[[y]],
#       nearest_PlotNumber = tree_sp$PlotNumber[nearest_idx],
#       nearest_MSP  = tree_sp[[x]][nearest_idx],
#       nearest_MCMT = tree_sp[[y]][nearest_idx],
#       delta_MSP  = gom_sp[[x]] - tree_sp[[x]][nearest_idx],
#       delta_MCMT = gom_sp[[y]] - tree_sp[[y]][nearest_idx],
#       mahalanobis_distance = mahal_dist,
#       outside_95 = mahal_dist > d95
#     )
#     
#     results[[sp]] <- out
#   }
#   
#   rbindlist(results, use.names = TRUE, fill = TRUE)
# }
```

```{r}
# gom.nearest <- compute_climate_nearest_treeplot(
#   tree_df = tree.climate2,
#   gom_df  = gom.climate,
#   x = "MSP",
#   y = "MCMT"
# )
# gom.species.mahal <- compute_species_mahalanobis(
#   tree_df = tree.climate2,
#   gom_df  = gom.climate,
#   x = "MSP",
#   y = "MCMT",
#   species_col = "species"
# )

```

```{r}

# gom.assessed <- fread("./clean_data/GoM_assessed_trials_2023_24.csv")
# gom.assessed.sf <- st_as_sf(gom.assessed, coords = c("Block_long", "Block_lat"), crs = 4326)
# 
# # Define performance levels and colors
# performance_levels <- c("Fail", "Poor", "Fair", "Good", "Excellent")
# performance_colors <- c(
#   "Excellent" = "purple",
#     "Good" = "green",
#     "Fair" = "yellow",
#    "Poor" = "orange",
#     "Fail" = "red"
# )
# 
# species_names <- c(
#   "Cw" = "Cw - Western redcedar",
#   "Fd" = "Fd - Douglas-fir",
#   "Lw" = "Lw - Western larch",
#   "Pw" = "Pw - Western white pine",
#   "Py" = "Py - Ponderosa pine"
#   )
# 
# plots <- list()
# 
# for (sp in unique(gom.sf$species)) {
#   gom_data <- gom.sf %>% filter(species == sp) %>% mutate(type = "GoM trial")
#   tree_data <- tree.sf %>% filter(species == sp) %>% mutate(type = "Natural stand")
#   assessed_data <- gom.assessed.sf %>%
#     filter(species == sp) %>%
#     mutate(performance = factor(performance, levels = performance_levels))
# 
#   # jitter assessed_data
#   jittered_coords <- st_coordinates(assessed_data) +
#     matrix(runif(nrow(assessed_data) * 2, -0.03, 0.03), ncol = 2)
#   assessed_data_jittered <- st_as_sf(
#     cbind(assessed_data, jittered_coords),
#     coords = c("X", "Y"),
#     crs = 4326
#   )
# 
# p <- ggplot() +
#   geom_sf(data = provmap_crop, fill = "grey90", color = "black") +
#   
#   geom_sf(data = gom_data, aes(shape = type), color = "black", size = 2, alpha = 0.8) +
#   geom_sf(data = assessed_data_jittered, aes(color = performance), size = 2, shape = 17, alpha = 0.9) +
#   geom_sf(data = tree_data, aes(shape = type), color = "darkblue", size = .8, alpha = 0.8) +
#   scale_shape_manual(
#     values = c("Natural stand" = 1, "GoM trial" = 2)
#   ) +
#   scale_color_manual(
#     values = performance_colors,
#     breaks = performance_levels
#   ) +
#   guides(
#     shape = guide_legend(order = 1),
#     color = guide_legend(order = 2)
#   ) +
#   coord_sf(
#     xlim = st_bbox(provmap_crop)[c("xmin", "xmax")],
#     ylim = st_bbox(provmap_crop)[c("ymin", "ymax")],
#     expand = FALSE
#   ) +
#   theme_minimal() +
#   theme(
#     axis.title.x = element_text(size = 8),
#     axis.title.y = element_text(size = 8),
#     axis.text = element_text(size = 8),
#     axis.ticks = element_line(color = "grey70", linewidth = 0.2),
#     panel.border = element_rect(color = "black", fill = NA, linewidth = 0.6),
#     plot.background = element_rect(color = "black", fill = NA, linewidth = 0.6),
#     plot.title = element_text(size = 10, face = "bold")
#   ) +
#   labs(
#     title = species_names[sp],
#     shape = "Source",
#     color = "GoM Assessed"
#   )
# 
# 
#   plots[[sp]] <- p
# }
# 
# 
# # Reorder plots so that Cw and Fd are last
# species_order <- c("Cw", "Fd", "Lw", "Pw", "Py")
# plots_ordered <- plots[species_order]
# 
# 
# 
# legend <- cowplot::get_legend(
#   plots_ordered[[1]] +
#     scale_shape_manual(
#       values = c("Natural stand" = 1, "GoM trial" = 2),
#       breaks = c("Natural stand", "GoM trial")
#     ) +
#     scale_color_manual(
#       values = performance_colors,
#       breaks = performance_levels
#     ) +
#     theme(
#       legend.position = "bottom",
#       legend.box = "horizontal",
#       legend.justification = "left",
#       legend.box.just = "left",
#       legend.text = element_text(hjust = 0),
#       legend.title = element_text(hjust = 0)
#     )
# )
# 
# 
# 
# plots_ordered_with_dist <- c(plots_ordered, list(distance_plot))
# 
# # Remove legends from all plots
# plots_nolegend <- lapply(plots_ordered_with_dist, function(x) x + theme(legend.position = "none"))
# 
# # Combine into 2 rows: first row has Lw, Pw, Py; second row has Cw, Fd
# combined_map_plot <- plot_grid(
#   plotlist = plots_nolegend,
#   nrow = 2,
#   labels = "AUTO",
#   label_size = 12,
#   label_fontface = "bold",
#   label_x = 0.03,   # inside border
#   label_y = 0.95,   # inside border
#   hjust = 0,
#   vjust = 0.5,
#   align = "hv",      # match widths
#   axis = "lr"       # align left/right edges
# )
# 
# #combined_map_plot <- combined_map_plot +
# #  theme(legend.position = "none")
# combined_map_plot
# 
# ggsave("./graphics/all_species_maps_new.jpg", combined_map_plot, width = 24.4, height = 16, units = "cm",dpi = 600)
# 
# 
# 
# # Add the shared legend underneath
# final_plot <- plot_grid(combined_map_plot, legend, ncol = 1, 
#                         rel_heights = c(1, 0.12),
#                   align = "hv",      # match widths
#                   axis = "lr" )
# final_plot
# #Save with specified width
#  ggsave("./graphics/all_species_maps_final.jpg", final_plot, width = 24.4, height = 16, units = "cm",dpi = 600)
# 
```



Density graphic of species across climate space
```{r}
# Load and filter data
library(data.table)
library(dplyr)
library(ggplot2)
library(ggdensity)

tree.climate2 <- fread("./clean_data/Tree_Plot_Climate.csv") %>% 
  filter(species %in% c("Cw","Fd","Lw","Pw","Py")) %>% 
  distinct()

gom.climate <- fread("./clean_data/GoM_trial_climate.csv") %>%
  filter(species %in% c("Cw","Fd","Lw","Pw","Py"))

WNA.climate <- fread("D:/OneDrive - Government of BC/CommonTables/BGC_climate/WNA_BGCv13_10km.csv") %>%
  filter(!is.na(MAT), !is.na(MAP), !is.na(MSP), !is.na(MWMT), 
         !is.na(CMD_an), !is.na(DD5_an))

gom.assessed <- fread("./clean_data/GoM_assessed_climate.csv") %>%
  filter(species %in% c("Cw","Fd","Lw","Pw","Py")) %>% filter(PERIOD == "1961_1990") %>% 
  arrange(species) %>%
  mutate(performance = factor(performance, 
                              levels = c("Fail", "Poor", "Fair", "Good", "Excellent"))) %>%
  distinct() %>%
  drop_na(performance)

# Define performance levels and colors
performance_levels <- c("Fail", "Poor", "Fair", "Good", "Excellent")
performance_colors <- c(
  "Fail" = "red",
  "Poor" = "orange",
  "Fair" = "yellow",
  "Good" = "green",
  "Excellent" = "purple"
)

species_names <- c(
  "Cw" = "Western redcedar",
  "Fd" = "Douglas-fir",
  "Lw" = "Western larch",
  "Pw" = "Western white pine",
  "Py" = "Ponderosa pine"
)

tree.climate2[, species_full := species_names[species]]
gom.assessed[, species_full := species_names[species]]
gom.climate[, species_full := species_names[species]]

# Plot
hdr.plot <- ggplot(tree.climate2, aes(x = DD5_an, y = MCMT)) +
  geom_hdr(data = tree.climate2, aes(x = DD5_an, y = MCMT), 
           probs = c(0.90, 0.75, 0.5)) +
  geom_point(data = gom.climate, aes(x = DD5_an, y = MCMT), 
             color = "grey20", alpha = 0.3, size = 1) +
  geom_point(
    data = gom.assessed,
    aes(x = DD5_an, y = MCMT, color = performance),
    size = 2, shape = 17, alpha = 0.9
  ) +
  scale_color_manual(
    values = performance_colors,
    drop = FALSE,
    guide = guide_legend(override.aes = list(shape = 17, size = 2))
  ) +
  coord_cartesian(xlim = c(250, 2000), clip = "off") +
  scale_x_continuous(limits = c(250, 2000))+ 
  facet_wrap(~ species_full, ncol = 3) +
  theme_minimal(base_size = 10) +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5),
    #strip.text = element_text(face = "bold", size = 12),
     strip.text = element_blank(),
  strip.background = element_blank(),
    legend.position = "bottom",
    legend.box = "vertical",
    axis.title.x = element_text(size = 8),
  axis.title.y = element_text(size = 8)

) +
  labs(x = "DD5_an", y = "MCMT")
hdr.plot
ggsave("./graphics/Species_Climate_Density_w_GoM_tall.png", 
       plot = hdr.plot, width = 8.6, height = 24.4, units = "cm")

library(patchwork)

species_labels <- c(
  "Cw - Western redcedar",
  "Fd - Douglas-fir",
  "Lw - Western larch",
  "Pw - Western white pine",
  "Py - Ponderosa pine"
)

label_df <- data.frame(
  species_full = species_labels,
  y = seq_along(species_labels)  # 1 to 5
)
label_plot <- ggplot(label_df, aes(x = 1, y = y, label = species_full)) +
  geom_text(hjust = .2, fontface = "bold", size = 3) + #
  scale_y_reverse() +
  coord_cartesian(clip = "off") +
  theme_void() +
  theme(
    plot.margin = margin(0, 0, 0, 0)  # tight to map
  )

combined <- (label_plot | combined_map_plot | hdr.plot) +
  plot_layout(
    widths = c(0.15, 0.425, 0.25),
    guides = "collect"
  ) &
  theme(legend.position = "bottom")



combined
ggsave("./graphics/Species_combined_tall.png", 
       plot = combined, width = 16.8, height = 24.4, units = "cm")

```