---
title: "Create Clean RESULTS data"
author: "Will MacKenzie & Hardy Griesbauer"
date: "2022-08-03"
output: html_document
---
The off-site trial data will be in 5 related tables.
- trial_site table is the master table and gives location information for an opening or trial that locates the trial and gives general features of the opening. For RESULTS data this is pulled from the Silviculture tables
- project_info is the metadata about the purpose and methods of a set of related trials
- trial_planting is the information on each offsite species planting occurring in a trial along with the qualitative assessment fields. For RESULTS data this is pulled from the PLANTING tables
- plot_site is for quantitative plots established in a trial
- plot_trees is the measurements of individual trees in a plot

The unique identifier tying these tables together is the Trial_ID. For RESULTS data this is the opening number.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(sf)
require(data.table)
require(tidyverse)
require(terra)
require(fasterize)
library(readr)
library(rmapshaper)
require(tictoc)

```

```{r selected species and CFRG tables}

offsite_spp <- c("FDI", "FDC", "LW", "PY", "PW", "CW","BA", "BG", "PA", "YC", "BN", "BP", "LS", "FD")

cfrg <- fread("./data/StockStands_v12_3.csv")
cfrg.offsite <-
  cfrg %>%
  mutate(Spp=toupper(Species)) %>% rename(BGC = ZoneSubzone) %>% select(BGC,Spp,Suitability) %>%  filter(Suitability >0 & Suitability <4) %>% distinct() %>% group_by(BGC, Spp) %>% filter(Suitability == min(Suitability)) %>% data.frame %>% 
  filter(!(Spp =='FD' & BGC == 'SBSdk'))
bgc <- vect("./spatial_files/BC_BGCv12_Published_Fixed_2.gpkg")
region <- vect("./spatial_files/Forest_Region_District.gpkg")
bgc_region <- vect("./spatial_files/BGC_Region_District.gpkg")

```

```{r import RESULTS data}
###---------This creates list of Openings with species of interest
silv <- st_read("./RESULTS_raw/WHSE_FOREST_VEGETATION_RSLT_FOREST_COVER_SILV_SVW.gpkg") %>% filter(!is.na(S_SPECIES_CODE_1))# %>%
 silv.dat <- silv %>% unite(c(BGC_ZONE_CODE, BGC_SUBZONE_CODE, BGC_VARIANT, BGC_PHASE), col = 'BGC', sep = "", na.rm = TRUE) %>%
   unite(c(BGC, BEC_SITE_SERIES), col = 'SS_NoSpace', na.rm = TRUE, sep = "/", remove = FALSE) %>%
  select(OPENING_ID, STANDARDS_UNIT_ID, S_SILV_LABEL, REFERENCE_YEAR, SITE_INDEX, BGC, SS_NoSpace, geom,  matches("SPECIES_CODE"))
# ###----To join with the planting layer for georeferencing
    silv_opening <- silv.dat %>% select(OPENING_ID, STANDARDS_UNIT_ID, S_SILV_LABEL, REFERENCE_YEAR, SITE_INDEX, BGC, SS_NoSpace, geom)

  silv.dat2 <- silv.dat %>%
     gather(-OPENING_ID, -STANDARDS_UNIT_ID, -geom, -S_SILV_LABEL, -REFERENCE_YEAR, -SITE_INDEX, -BGC, -SS_NoSpace, key = "species_col", value = "Spp") %>% filter(Spp %in% offsite_spp)

   st_write(silv.dat2,"./RESULTS_raw/silv_w_spp_of_interest.gpkg", delete_layer = TRUE)
# # ### Bring into QGIS and spatial join to BGC
 silv.dat <- st_read("./RESULTS_raw/silv_w_spp_of_interest_bgc.gpkg")
 silv.dat <- silv.dat %>% mutate(SppVar = Spp, BGC = BGC_2) %>% 
mutate(Spp = recode (Spp, 'FDI' = "FD", 'FDC' = "FD"))
offsite_silv <-  anti_join(silv.dat, cfrg.offsite, by=c("BGC" , "Spp"))
st_write(offsite_silv, "./clean_data/silv_openings_w_offsite.gpkg", delete_layer = TRUE)

###----bring in planting layer as the master spatial file to join Silv info to.
plant <- st_read("./RESULTS_raw/WHSE_FOREST_VEGETATION_RSLT_PLANTING_SVW.gpkg")# %>%
 plant.dat <- plant %>%  filter(!is.na(SILV_TREE_SPECIES_CODE)) %>%  dplyr::select(ACTIVITY_TREATMENT_UNIT_ID,
                OPENING_ID,
                MAP_LABEL,
                SILV_BASE_CODE,
                SILV_TECHNIQUE_CODE,
                SILV_METHOD_CODE,
                ATU_COMPLETION_DATE, # use to calculate age
                ACTUAL_TREATMENT_AREA,
                SILV_TREE_SPECIES_CODE,
                NUMBER_PLANTED, # number planted
                SEEDLOT_NUMBER,
                GENETIC_CLASS_CODE) %>%
   filter(SILV_TREE_SPECIES_CODE %in% offsite_spp) %>% mutate(SppVar = SILV_TREE_SPECIES_CODE) %>%  mutate(Spp = recode(SppVar, "FDI" = "FD", "FDC" = "FD")) %>%
data.frame %>% select(-geom)
#
plant.dat2 <- left_join(plant.dat, silv_opening, by = "OPENING_ID") %>% st_as_sf %>% filter(!st_is_empty(.))
plant_silv <- as.data.frame(plant.dat2)
silv_noplant <- anti_join(offsite_silv, plant_silv, by=c("OPENING_ID", "Spp")) %>% st_as_sf %>% filter(!st_is_empty(.))## where off-site silv records but no plant record
silv_noplant <- as.data.frame(silv_noplant)
plant.dat.all <- trials.merge(plant_silv, silv_noplant)
st_write(plant.dat.all,"./RESULTS_raw/planting_w_spp_of_interest2.gpkg", delete_layer= TRUE)
 ##---------- Bring  above into QGIS and do spatial join to latest BGC
```


```{r adjust combined data for use in offsite scripts}
plant.dat3 <- st_read("./RESULTS_raw/planting_w_spp_of_interest_bgc.gpkg") %>% mutate(BGC = BGC_2)

offsite_plant <-  anti_join(plant.dat3, cfrg.offsite, by=c("BGC" , "Spp")) ### species not in BGC from CFRG
offsite_plant$PlantingYear <- as.numeric(substr(offsite_plant$ATU_COM, 1, 4))
offsite_plant$PlantingDate <- as.character(substr(offsite_plant$ATU_COM, 1, 10))
offsite_plant$PlantingSeason <- as.numeric(substr(offsite_plant$ATU_COM, 6, 7))
st_write(offsite_plant, "./clean_data/RESULTS_offsite.gpkg", delete_layer = TRUE) 
 
offsite_plant2 <- vect(offsite_plant)
y <- centroids(offsite_plant2 , TRUE) %>% st_as_sf
st_write(y, "./RESULTS_raw/RESULTS_offsite_centroids.gpkg", delete_dsn = TRUE) 


 ##---------- Bring above file into QGIS and do spatial join to latest BGC - this should be scripted but need a fast method in r
y <- st_read("./RESULTS_raw/RESULTS_offsite_centroids_bgc.gpkg") %>% mutate(BGC = BGC_3)
offsite_plant3 <-  anti_join(y, cfrg.offsite, by=c("BGC" , "Spp"))%>% distinct(OPENING_ID, Spp, .keep_all = TRUE) %>% 
  select(-BGC_2, -BGC_3)
offsite_good <- offsite_plant3 %>% filter(!is.na(PlantingYear))  %>% as.data.frame%>% select(OPENING_ID, PlantingYear, PlantingSeason, PlantingDate)
offsite_bad <- offsite_plant3 %>% filter(is.na(PlantingYear)) %>% as.data.frame %>% select( -PlantingYear, -PlantingSeason, -PlantingDate) 
offsite_update <- left_join(offsite_bad, offsite_good, by = "OPENING_ID") %>% distinct
offsite_good <- offsite_plant3 %>% filter(!is.na(PlantingYear)) %>% data.frame

offsite_plant3 <- trials.merge(offsite_good, offsite_update) %>% st_as_sf()

st_write(offsite_plant3, "./clean_data/RESULTS_offsite_centroids_cleaned.gpkg", delete_dsn = TRUE) 
offsite_planttable <- st_transform(offsite_plant3, 4326) %>%  
  mutate(Longitude = sf::st_coordinates(.)[,1], Latitude = sf::st_coordinates(.)[,2]) %>% sf::st_set_geometry(NULL) %>% as.data.frame() %>% 
 rename('Trial_ID' = OPENING_ID, 'SpeciesPlanted' = SILV_TREE_SPECIES_CODE, ATU_ID = ACTIVITY_TREATMENT_UNIT_ID, 
        SppComposition_Label = S_SILV_LABEL,  Seedlots = SEEDLOT_NUMBER, StockType = SILV_METHOD_CODE, 
        Seed_Class = GENETIC_CLASS_CODE, SiteIndex = SITE_INDEX) %>%  
 mutate(Longitude = as.numeric(Longitude), Latitude = as.numeric(Latitude), Trial_Type = "Operational", 
        Project_Name = "RESULTS",  BGCv12 = BGC,
        ATU_ID = as.character(ATU_ID), Trial_ID = as.character(Trial_ID)) %>%   
select(-ATU_COMPLETION_DATE, -REFERENCE_YEAR)

fwrite(offsite_planttable, "./clean_data/RESULTS_offsite_points.csv" )
```


