---
title: "Qualitative Species Performance Ratings of AMAT trials"
author: "William H MacKenzie"
date: "20/09/2024"
format:
  typst:
    toc: false
    toc-depth: 2
    toc-title: Contents
    section-numbering: 1.1.1
    columns: 1
    fig-width: 7
    fig-height: 10
    fig-caption: true
    fig-dpi: 600

editor: visual
execute:
  echo: false
  error: false
  warning: false
  message: false
  fig.align: 'left'
  fig.cap: true
  fig.pos: H
  dev: pdf
  fig.ext: pdf
  cache: false
  fig.retina: 2
  dpi: 600
  fig.path: "./figures/"
---

Qualitative ratings of performance are generated from survival and height data collected from long-term trials. The best performing Stock_IDs for each species is identified by selecting the Stock_IDs with max(sum of heights). The single metric balances survival and growth. A mean performance (MP) is calculated by dividing the sum of heights by the number of trees planted. The rating is used to classify the performance of the Stock_IDs into five qualitative performance categories:

-   Excellent - \>4.0 m MP and survival \>75%,

-   Good - \>2 m MP and survival \>66%),

-   Fair - \>1.5m MP or survival \>75%,

-   Poor - survival \>30%,

-   Fail - survival \<30%.

The best performing Stock_IDs are mapped to show the distribution of the qualitative rating for species across the trial sites.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
require(sf)
require(data.table)
require(tidyverse)
require(terra)
require(fasterize)
library(readr)
library(rmapshaper)
require(tictoc)
require(ggplot2)
require(forcats)
require(knitr)
require(data.table)
require(vegan)
require(ggpubr)
require(ggrepel)
source("./_functions/doc_theme_pem.R")

```

```{r import AMAT trial data and clean, echo=FALSE}
dead = c(".", "..", "", "0", "9999", "x")
data1 <- readxl::read_excel("./data/AMAT_MASTERDATA_ver_14.xlsx", sheet = "growth")
data1 <- data1 %>% select(Site, Sp, SLnum, Rep, HT10) %>%
  mutate(SLnum = as.factor(SLnum)) %>%
  mutate(Stock_ID = paste0(Sp, "-", SLnum)) %>%
  mutate(HT10 = ifelse(HT10 %in% dead, 0, HT10)) %>%
  #filter(!HT10 %in% dead) %>%
  mutate(HT10 = as.numeric(HT10)) %>%
  mutate(height = ifelse(HT10 == 0, NA, HT10/100)) %>%
  mutate(height = round(height, 2))
 seedlot.info <- readxl::read_excel("./data/AMAT_MASTERDATA_ver_14.xlsx", 
                            sheet = "seedlots") %>%
  mutate(SLnum = as.factor(SLnum)) %>% 
   select(SLnum, Seedlot_label)
 data1 <- left_join(data1, seedlot.info) %>% select(Site, Sp, Stock_ID, SLnum, Seedlot_label, everything())
  #mutate_if(is.character, as.factor) %>%%>% droplevels()# %>% rowid_to_column("ID")
fwrite(data1, "./clean_data/AMAT_10year.csv" )

```

```{r summarize AMAT and select best performing Stock_IDs, echo=FALSE, warning=FALSE}
data1 <- fread("./clean_data/AMAT_10year.csv")
trees <- data1 %>% 
  select(Site, Sp, Stock_ID, Seedlot_label,HT10, height) %>% 
  group_by(Site, Sp, Stock_ID, Seedlot_label) %>% 
 #   summarise(planted = n(), survived = sum(!is.na(height))) %>% 
  summarise(mean.ht = mean(height, na.rm = TRUE), sum.ht = sum(height, na.rm = TRUE),
            best.ht = max(height, na.rm = TRUE), planted = n(), survived = sum(!is.na(height))) %>% 
  mutate(survival = round((survived/planted*100),0)) %>% 
  mutate(best.ht = replace(best.ht, best.ht == '-Inf', 0)) %>% 
  mutate(mean.ht = replace(mean.ht, is.na(mean.ht), 0)) %>%
  ungroup %>%
  group_by(Site, Sp) %>%
  mutate(best.perform = max(sum.ht)) %>% ungroup %>% 
  data.frame %>%
  mutate(mean.perform = round((best.perform/planted),2)) %>% 
  mutate(rating = ifelse(mean.perform > 4 & survival > 75, "1_Excellent",
                  ifelse(mean.perform > 2.0 & survival > 66, "2_Good",
                  ifelse(mean.perform > 1.5 | survival > 75, "3_Fair",
                  ifelse(survival >30,
                         "4_Poor","5_Fail")))))
sites <- readxl::read_excel("./data/AMAT_MASTERDATA_ver_14.xlsx", 
                            sheet = "sites") %>% 
  select(Site, SiteName, BECvar, LONG_S, LAT_S) %>% rename(longitude = LONG_S, latitude = LAT_S) %>%
  filter(!Site %in% c("0", "9999", "x")) %>% 
  mutate(Site = as.factor(Site))
trials.all <- trees %>% select(Site, Sp, Seedlot_label, mean.perform, rating) %>% 
  left_join(sites) %>% arrange(Sp)
fwrite(trials.all, "./clean_data/AMAT_10year_all_assessed.csv")
best.Stock_ID <- trials.all  %>% group_by(Site, Sp) %>% slice_max(mean.perform, n=1, with_ties = FALSE) %>% ungroup %>% distinct %>% data.frame
xx <- trees %>% filter(Sp ==  'Lw')
fwrite(best.Stock_ID, "./clean_data/AMAT_10year_best_Stock_ID.csv")
```

```{r map function, echo=FALSE}
trial.map <- function(dat = trial.dat, spp, basemap = map){
trial.dat1 <- trial.dat %>% mutate_if(is.character, as.factor) %>% filter(Sp %in% spp)
trial.loc <- st_as_sf(trial.dat1, coords=c("longitude", "latitude"), crs = 4326)
trial.coords <- st_transform(trial.loc, crs= st_crs(basemap)) %>% st_coordinates() %>% as.data.frame
trial.loc2 <- as.data.frame(trial.loc) %>% cbind(trial.coords) # convert back to data.frame for ggplot
map_intense <- ggplot() +
  geom_sf(data = basemap, color = "grey 70", fill = "grey 95", size = 1)  +
  theme_light()+
geom_point(data = trial.loc2, size = 2, aes(X,Y, colour = rating, shape = rating), show.legend = TRUE) +
  scale_colour_manual(drop = FALSE, values = c('purple', 'green', 'orange', 'red', 'black'))+
  scale_shape_manual(drop=FALSE, values = c(17, 16, 15, 25, 4), )+
  ggrepel::geom_text_repel(data = trial.loc2, mapping = aes(X, Y, label = SiteName),
                  box.padding   = 0.1,
                  point.padding = 0.1,
                  segment.color = 'black',
                  min.segment.length = 0,
                  direction = 'both',
                  size = 2,
                  max.overlaps = 30)+
  theme(legend.position="bottom") +
  labs(title = paste0(spp, " trial locations"), x="longitude",  y="latitude")
#map_intense
print(map_intense)  
ggsave(paste0("./maps/", spp, " AMAT performance.pdf"), width = 7, height = 10, units = "in")
#return(map_intense)
}


```

## AMAT species trials maps

Maps for each tree species showing the performance rating for the best performing Stock_ID at each AMAT trial location.

```{r map distribution of best performing Stock_ID ratings, echo=FALSE, error = FALSE, warning=FALSE, message=FALSE}

trial.dat <- fread("./clean_data/AMAT_10year_best_Stock_ID.csv")
map <- st_read("D:/CommonTables/BC_AB_US_Shp/WNA_State_Boundaries.gpkg", layer = "WNA_State_Boundaries", quiet = TRUE)
st_crs(map) <- 3005
spp.codes = "At"
spp.codes = str_sort(unique(trial.dat$Sp))
for (spp in spp.codes){
   trial.map(trial.dat, spp, map)
}
#dist.map(trial.dat, "Py", map)
```
